Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors

### 2021-10-06 DOI_Review/Manuscript.rmd
refs ~done~

While some debate still exists and exact methods vary, it is largely accepted that community-based conservation interventions are a viable method for seeding self-spreading, -sustaining, and ethical protections on natural resources [TURNER 2017; BEREKES 2007; ADAMS2004]. A major question in implementing these projects is if a specific intervention will indeed catch-on in a population and if so, how quickly [Radeloff 2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [Mangubhai 2011; Visconti 2019 ].

# Diffusion Theory
Inducing and/or speeding up the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [ROGERS 2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, the core of diffusion theory is the notion that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently.

This assumption about human decision-making has been validated repeatedly in cultures around the world; individuals do not act as cost-benefit analysts, in which they adapt their behaviors based on the environment alone [HENRICH et al 2001]. Our dependence on social learning to make decisions about behavioral adoption has profound and specific implications for patterns of behavioral adoption at the population level [SMALDINO 2018; another cite??]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a behavior as a function of time [HOPPITT 2013; ROGERS 2010; RYAN & GROSS 1950].

Behavioral spread through imitation produces a sigmoidal “S” shaped diffusion curve, which is observed in the vast majority of empirical studies on behavioral adoption [HENRICH 2001]. This “S” shaped curve is produced regardless of the social learning mechanism employed by a population and can even be observed when a portion of the population relies on individual learning [HENRICH 2001]. Behavioral adoption thus exhibits equifinality, where many different underlying mechanisms may produce similar or even identical patterns over time [BARRETT 2019].

@@

Individuals tend to place more weight on imitation when making decisions in which the outcomes can not be well predicted independently [PERREAULT 2012]. This phenomenon is particularly salient under variable, rapidly changing environments [BOYD & RICHERSON 1988; Richerson & Boyd 2008]. It follows then, that the adoption of novel conservation behaviors by a community is likely to be achieved through social imitation (i.e. diffusion), especially given worsening climatic instability and regime changes resulting from widespread biodiversity loss.

Indeed, the sigmoidal pattern of behavioral adoption characteristic to diffusion has been observed in over 80% of case studies on community-based conservation interventions worldwide [MASCIA & MILLS 2018; MILLS et al 2019]. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the adoption of community based conservation interventions is considerably attenuated [PIERTI et al 2009]. With the recognition that the patterns of uptake for community-based conservation practices are directly linked to behavioral imitation, researchers have begun suggesting diffusion theory as a lens for understanding and implementing individual interventions [ABERNETHY et al 2014; MBARU & BARNES 2017; NIEMIC et al 2021].

Meta-level studies have recently come to the same conclusion. For example, in a review of 22 conservation interventions referenced above, the majority of  initiatives were adopted by less than 30% of potential adopters, leading the authors to point to diffusion theory as an avenue for increasing the proportion of adopters [MILLS et al 2019]. And a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices further identified diffusion theory as useful for guiding the development of conservation interventions [MAHAJAN et al 2020].

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [SMALDINO & JONES 2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral uptake [CAVALLI-SFORZA and FELDMAN 1981; STRANG 1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [FARZANEH et al 2019;SMALDINO & JONES 2020].

@@

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, &  SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [HUPPERT & KATRIEL 2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [Grinsztajn 2020; CARPENTER 2018]. Below, we use a minimal agent-based model of conservation adoption contingent on behavior-environment feedbacks to simulate data on the long-term adoption dynamics of a community-based conservation intervention. Then, using just the first 30% of the data, we estimate the rate of intervention spread, rate of dropout, and time after dropout before agents are available to adopt again. We use these estimated parameters to predict the overall patterns of adoption over all time steps and compare our observed to predicted values.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under both unbiased and success-biased social learning. Success-biased social learning is contingent on behavior-environment feedbacks. Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [FERRARO 2011]. We establish a degraded landscape/seascape (from herein referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first timestep, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and individuals are rewarded for abstaining from harvesting there. In each subsequent timestep, agents either choose to enroll in the conservation program, receive their reward, and continue to harvest just from the remaining 80% of the unprotected landscape, or they choose to harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior [BALDINI 2013, BARRETT 2019]. The protected and unprotected resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area.

@@

With data on conservation behavior adoption for both learning mechanisms in-hand, we fit the SIRS model described above to the first 30% of observations. We use probabilistic sampling to infer the parameters $\beta$, $\gamma$, and $\chi$, for the full Bayesian equation below where $\text{Y}_{t}$ represents the total number of individuals who are enrolled in the community-based intervention at any given time:

@@

As demonstrated above, the standard SIRS diffusion model can reliably capture the long-term patterns of behavioral adoption for community-based conservation interventions, even when adoption is contingent on feedbacks with ecological processes. This model, while accurately capturing the pattern of spread, is unable to provide insight as to the mechanism of spread or causality in the system. As shown above and in existing literature, the social transmission mechanism, or mix of mechanisms, employed by a population can dramatically affect the overall rates of adoption for a given behavior [SMALDINO et al 2017; JEFF&MONIQUE2018; WARING2021]. Therefore, while we agree that diffusion theory can be productively applied to community-based conservation, we point to theories of social learning and cultural evolution as a better alternative for diagnosing and planning interventions.

Understanding the causal mechanisms behind the patterns of behavioral transmission is necessary to better delineate and predict how conservation will affect both people and the environment [CHENG et al 2020]. Researchers are increasingly recognizing that models of human-caused environmental change must explicitly incorporate dynamic social processes [TURNER et al 2021]. Specifically, in modeling social-ecological systems, either using analytical approaches or simulation, processes should be based on realistic human cognitions which allow actors to modify their behaviors in response to new information [MEYFROIDT 2013]. Modeling human behavior in the context of community-based conservation is thus going to require a complex dynamic systems approach in which behaviors and the biophysical environment coevolve [SCHILL et al 2019].

We argue that when prediction is the goal, analytical or statistical approaches like the SIRS model presented above are always preferable to simulation based approaches [EPSTEIN 2008; TREDENNINCK et al 2021]. The practice of applying statistical frameworks of parameter estimation to analytical process models, and in particular to ordinary differential equations, is still relatively new and has been embraced mostly by ecology [HEFLEY 2017]. Extending the SIRS diffusion model to include a term which represents changes in the environment over time and the effect of resource levels on adoption rates may more accurately describe the system of adoption in community-based conservation interventions. An example of such an extension is presented below:

@@

While process models may be preferable for predicting adoption outcomes of conservation interventions, they might not always be possible. In complex systems, these models can quickly require many parameters in order to accurately account for all moving parts and thus are prone to issues of identifiability in parameter estimation  [LAUBMEIER 2020]. Further, complex social-ecological systems have properties which are inherently stochastic and may be impossible to model analytically. For example, coastal fisheries have been shown to approach mathematical chaos in the number and proportions of species present [WILSON et al 1994].

In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for tipping points and emergence, two key features of the complex systems which can not be accounted for with process models such the SIRS or SIRS extension models presented above [REYERS et al 2018; RAILSBACK & GRIMM; WILENSKY & RAND]. Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically  [LIU et al 2007].

>>>>

While some debate still exists and exact methods vary, it is largely accepted that community-based conservation interventions are a viable method for seeding self-spreading, -sustaining, and ethical protections on natural resources [@turner_global_2007; @berkes_community-based_2007; @adams_biodiversity_2004]. A major question in implementing these projects is if a specific intervention will indeed catch-on in a population and if so, how quickly [@radeloff_hot_2013]? Predicting the final proportion of a population that will adopt an intervention is critical, as many interventions require buy-in from nearly all community members to be effective [@mangubhai_not_2011; @visconti_protected_2019].

# Diffusion Theory
Inducing and/or speeding up the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, the core of diffusion theory is the notion that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently.

This assumption about human decision-making has been validated repeatedly in cultures around the world; individuals do not act as cost-benefit analysts, in which they adapt their behaviors based on the environment alone [@henrich_search_2001]. Our dependence on social learning to make decisions about behavioral adoption has profound and specific implications for patterns of behavioral adoption at the population level [@smaldino_adoption_2017]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a behavior as a function of time [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950].

Behavioral spread through imitation produces a sigmoidal “S” shaped diffusion curve, which is observed in the vast majority of empirical studies on behavioral adoption [@henrich_cultural_2001]. This “S” shaped curve is produced regardless of the social learning mechanism employed by a population and can even be observed when a portion of the population relies on individual learning [@henrich_cultural_2001]. Behavioral adoption thus exhibits equifinality, where many different underlying mechanisms may produce similar or even identical patterns over time [@barrett_equifinality_2019].

@@

Individuals tend to place more weight on imitation when making decisions in which the outcomes can not be well predicted independently [@perreault_bayesian_2012]. This phenomenon is particularly salient under variable, rapidly changing environments [@boyd_culture_1988; @richerson_not_2008]. It follows then, that the adoption of novel conservation behaviors by a community is likely to be achieved through social imitation (i.e. diffusion), especially given worsening climatic instability and regime changes resulting from widespread biodiversity loss.

Indeed, the sigmoidal pattern of behavioral adoption characteristic to diffusion has been observed in over 80% of case studies on community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the adoption of community based conservation interventions is considerably attenuated [@pietri_information_2009-1]. With the recognition that the patterns of uptake for community-based conservation practices are directly linked to behavioral imitation, researchers have begun suggesting diffusion theory as a lens for understanding and implementing individual interventions [@abernethy_two_2014; @mbaru_key_2017; @niemiec_efficacy-based_2021].

Meta-level studies have recently come to the same conclusion. For example, in a review of 22 conservation interventions referenced above, the majority of  initiatives were adopted by less than 30% of potential adopters, leading the authors to point to diffusion theory as an avenue for increasing the proportion of adopters [@mills_how_2019]. And a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices further identified diffusion theory as useful for guiding the development of conservation interventions [@mahajan_theorybased_2020].

@@

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral uptake [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. In this instantiation, $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

@@

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, &  SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017;  @carpenter_predator-prey_2018]. Below, we use a minimal agent-based model of conservation adoption contingent on behavior-environment feedbacks to simulate data on the long-term adoption dynamics of a community-based conservation intervention. Then, using just the first 30% of the data, we estimate the rate of intervention spread, rate of dropout, and time after dropout before agents are available to adopt again. We use these estimated parameters to predict the overall patterns of adoption over all time steps and compare our observed to predicted values.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under both unbiased and success-biased social learning. Success-biased social learning is contingent on behavior-environment feedbacks. Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (from herein referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first timestep, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and individuals are rewarded for abstaining from harvesting there. In each subsequent timestep, agents either choose to enroll in the conservation program, receive their reward, and continue to harvest just from the remaining 80% of the unprotected landscape, or they choose to harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior [@baldini_success-biased_2012; @baldini_two_2013, @barrett_equifinality_2019]. The protected and unprotected resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area.

@@

With data on conservation behavior adoption for both learning mechanisms in-hand, we fit the SIRS model described above to the first 30% of observations. We use probabilistic sampling to infer the parameters $\beta$, $\gamma$, and $\chi$, for the full Bayesian equation below where $\text{Y}_{t}$ represents the total number of individuals who are enrolled in the community-based intervention at any given time. Here, "infected" individuals are those that are enrolled in the conservation intervention.

@@

As demonstrated above, the standard SIRS diffusion model can reliably capture the long-term patterns of behavioral adoption for community-based conservation interventions, even when adoption is contingent on feedbacks with ecological processes. This model, while accurately capturing the pattern of spread, is unable to provide insight as to the mechanism of spread or causality in the system. As shown above and in existing literature, the social transmission mechanism, or mix of mechanisms, employed by a population can dramatically affect the overall rates of adoption for a given behavior [@smaldino_adoption_2017; @andrews_cultural_2018; @tam_measuring_2021]. Therefore, while we agree that diffusion theory can be productively applied to community-based conservation, we point to theories of social learning and cultural evolution as a better alternative for diagnosing and planning interventions.

Understanding the causal mechanisms behind the patterns of behavioral transmission is necessary to better delineate and predict how conservation will affect both people and the environment [@cheng_strengthen_2020]. Researchers are increasingly recognizing that models of human-caused environmental change must explicitly incorporate dynamic social processes [@turner_land-useland-cover_2021]. Specifically, in modeling social-ecological systems, either using analytical approaches or simulation, processes should be based on realistic human cognitions which allow actors to modify their behaviors in response to new information [@meyfroidt_environmental_2013]. Modeling human behavior in the context of community-based conservation is thus going to require a complex dynamic systems approach in which behaviors and the biophysical environment coevolve [@schill_more_2019].

We argue that when prediction is the goal, analytical or statistical approaches like the SIRS model presented above are always preferable to simulation based approaches [@epstein_why_2008; @tredennick_practical_2021]. The practice of applying statistical frameworks of parameter estimation to analytical process models, and in particular to ordinary differential equations, is still relatively new and has been embraced mostly by ecology [@hefley_when_2017]. Extending the SIRS diffusion model to include a term which represents changes in the environment over time and the effect of resource levels on adoption rates may more accurately describe the system of adoption in community-based conservation interventions. An example of such an extension is presented below where we include another equation for the change in a resource pool over time $\frac{dE}{dt}$. We also must include three additional parameters: $\phi$. $\rho$, and $\theta$, which represent the effect of resource level on enrollment (infection), the effect of harvest on the resource level, and the regeneration rate of the resource respectively.

@@

While process models may be preferable for predicting adoption outcomes of conservation interventions, they might not always be possible. In complex systems, these models can quickly require many parameters in order to accurately account for all moving parts and thus are prone to issues of identifiability in parameter estimation [@laubmeier_ecological_2020]. Further, complex social-ecological systems have properties which are inherently stochastic and may be impossible to model analytically. For example, coastal fisheries have been shown to approach mathematical chaos in the number and proportions of species present [@wilson_chaos_1994].

In very complex cases such as these, we argue that agent-based simulations are the preferred approach as they allow for tipping points and emergence, two key features of the complex systems which can not be accounted for with process models such the SIRS or SIRS extension models presented above [@reyers_social-ecological_2018; @railsback_agent-based_2012; @wilensky_introduction_2015]. Coupled human and natural systems commonly exhibit dynamics with thresholds, resilience, and heterogeneity not present in social or natural systems in isolation and therefore are difficult to model analytically  [@liu_complexity_2007].
